Mergeable Types

### 2017-02-27 paper/model.tex
Model.

A significant advantage of mergeable types is their composability. As
demonstrated by \C{MList}, the merge semantics of a composed mergeable
type is a composition of merge semantics of the individual types.  For
instance, a mergeable list of integers can be obtained by
instantiating \C{MList} for an \C{Int} type, where \C{Int.merge} is
same as \C{Counter.merge} (Sec.~\ref{sec:intro}). 
An alternative instantiation of \C{MList} is for a quantity type
(\C{Qty}) that, like \C{Int}, supports \C{add} and \C{subtract}
operations, but defines its \C{merge} semantics a little differently:
\begin{ocaml}
let merge lca v1 v2 = if v1 = lca then v2 else (if v2 = lca then v1 else max v1 v2)
\end{ocaml}
\C{MList(Qty)}, like \C{MList(Int)}, is a list of integers, and has
the same semantics under a sequential execution. Their differences
however manifest under a concurrent execution, where both adopt
different methods to reconcile concurrent updates.  is given
below.

% The mergeable list data type demonstrates a common pattern of writing
% \C{merge} functions applicable to a range of functional data
% structures. For instance, a binary tree (with rotations) can be made
% mergeable in the same way as list, i.e., by computing edit sequences
% (\C{edit\_seq}) and then their transformations
% (\C{op\_transform})\footnote{
%   Like Wagner-Fischer, established algorithms exist for computing tree
%   edit distances~\cite{tree-diff}. An alternative naive method to
%   merge a pair of trees is by merging the lists obtained from their
%   in-order traversal, followed by reconstructing the tree.
% }.

{\bf Shopping List}. A non-trivial demonstration of composability of
mergeable types is a shopping list application. The application
allows its users to collaboratively build a grocery shopping
list (listing items in decreasing ), by adding new items with a
specified quantity, deleting existing items, and increasing or
decreasing their quantity. An item (\C{Item.t}) is represented as a
tuple of its name (a string) and the quantity (A \C{Qty.t}). Merge
function for \C{Item.t} merges the quantities of items (using
\C{Qty.merge}) with same names. Items with different names are
considered distinct. A shopping list is a mergeable list of items
(i.e., \C{MList(Item)}). \C{MList(Item).merge} automatically lifts the
merge semantics for items to merge semantics for lists of items.
Fig.~\ref{fig:shopping-list} illustrates. Here, two users, Alice and
Bob, concurrently edit a shopping list whose initial contents are
as milk and eggs. While both simultaneously update the quantity of
eggs, Bob also removes milk and inserts candy. The merged shopping
list contains eggs and candy, where the quantity of eggs is obtained
by merging Alice's and Bob's quantity as per \C{Qty.merge} semantics.

>>>>

A significant advantage of mergeable types is their composability. As
demonstrated by \C{MList}, the merge semantics of a composed mergeable
type is a composition of merge semantics of the individual types.  For
instance, a mergeable list of integers can be obtained by
instantiating \C{MList} for an \C{Int} type, where \C{Int.merge} is
same as \C{Counter.merge} (Sec.~\ref{sec:intro}). 
An alternative instantiation of \C{MList} is for a quantity type
(\C{Qty}) that, like \C{Int}, supports \C{add} and \C{subtract}
operations, but defines its \C{merge} semantics a little differently:
\begin{ocaml}
let merge lca v1 v2 = if v1 = lca then v2 else (if v2 = lca then v1 else max v1 v2)
\end{ocaml}
\C{MList(Qty)}, like \C{MList(Int)}, is a list of integers, and has
the same semantics under a sequential execution. Their differences
however manifest under a concurrent execution, where both adopt
different methods to reconcile concurrent updates. 
% The mergeable list data type demonstrates a common pattern of writing
% \C{merge} functions applicable to a range of functional data
% structures. For instance, a binary tree (with rotations) can be made
% mergeable in the same way as list, i.e., by computing edit sequences
% (\C{edit\_seq}) and then their transformations
% (\C{op\_transform})\footnote{
%   Like Wagner-Fischer, established algorithms exist for computing tree
%   edit distances~\cite{tree-diff}. An alternative naive method to
%   merge a pair of trees is by merging the lists obtained from their
%   in-order traversal, followed by reconstructing the tree.
% }.

{\bf Shopping List}. A non-trivial demonstration of composability of
mergeable types is a shopping list application. The application allows
its users to collaboratively build a grocery shopping list (listing
items in the order of their priority), by adding new items with a
specified quantity, deleting existing items, and increasing or
decreasing their quantity. An item (\C{Item.t}) is represented as a
tuple of its name (a string) and the quantity (A \C{Qty.t}). Merge
function for \C{Item.t} merges the quantities of items (using
\C{Qty.merge}) with same names. Items with different names are
considered distinct. A shopping list is a mergeable list of items
(i.e., \C{MList(Item)}). \C{MList(Item).merge} automatically lifts the
merge semantics for items to merge semantics for lists of items.
Fig.~\ref{fig:shopping-list} illustrates. Here, two users, Alice and
Bob, concurrently edit a shopping list whose initial contents are as
milk and eggs. While both simultaneously update the quantity of eggs,
Bob also removes milk and inserts candy. The merged shopping list
contains eggs and candy, where the quantity of eggs is obtained by
merging Alice's and Bob's quantity as per \C{Qty.merge} semantics.
